Data Imputation as a Query Optimization

### 2016-12-10 paper/algorithm.tex

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric which combines the cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.

@@

$\text{Loss}(Q)$ is computed as follows.
For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered. First, all filters are pushed to the leaves of the query tree, immediately after the scans. Joins are performed after filtering, and only left-deep plans are considered. Any group-by/aggregate will be performed after the final join. Finally, projections are placed at the root of the query tree. The space of query plans is similar to that considered by System R\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

\subsection{Query planning}
The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregator function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.

>>>>

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.

@@

\todo{Check loss calculation w/ Micah}
$\text{Loss}(Q)$ is computed as follows.
For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered. First, all filters are pushed to the leaves of the query tree, immediately after the scans. Joins are performed after filtering, and only left-deep plans are considered. Any group-by/aggregate will be performed after the final join. Finally, projections are placed at the root of the query tree. The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

\subsection{Query planning}
The input to our query planner is a tuple $(T, \Phi, \Psi, P, G, A)$: a set of tables $T$, a set of filter predicates $\phi_t, t \in T$, a set of join predicates $\psi_(t_1, t_2), t_1, t_2 \in T$, a set of attributes $P$, and an optional set of attributes $G$ and aggregation function $A \in \{\text{Max}, \text{Min}, \text{Sum}, \text{Avg}, \text{Count}\}$.
